Making Sequential Programs Deterministic

### 2019-04-06 conclusion.tex

\todo{TODO}

We designed a type system that helps programmers express determinism specifications
in sequential programs. To the best of our knowledge, ours is the
first compile time verification approach addressing the problem of nondeterminism.
We implemented our type system in Java, applied it to real world 
software and found it to be effective in finding previously unknown determinism bugs. 
\TheDeterminismChecker found errors that the
Randoop developers had missed, despite spending over 100 hours on the
problem of nondeterminism. We compared our tool with existing state of the art flake test detectors
and found that \TheDeterminismChecker found all the bugs arising due to nondeterminism
in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.

>>>>

We designed a type system that helps programmers express determinism specifications
in sequential programs. 
We implemented our type system in Java, applied it to real world 
software and found it to be effective in finding previously unknown determinism bugs. 
\TheDeterminismChecker found errors that the
Randoop developers had missed, despite spending over 100 hours on the
problem of nondeterminism. We compared our tool with existing state of the art flake test detectors
and found that \theDeterminismChecker found all the bugs arising due to nondeterminism
in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.
To the best of our knowledge, ours is the
first compile time verification approach addressing the problem of nondeterminism.

### 2019-04-06 paper.tex

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
When a program is nondeterministic, it is difficult to test and debug.
Nondeterminism can occur even in sequential programs, for example as a
result of iterating over the elements of a hash table.

We have created a type system for determinism.  A type checker warns if a
program is nondeterministic or verifies that the program is deterministic.
The key ideas in the type system are type qualifiers for nondeterminism,
order-nondeterminism, and determinism; type well-formedness rules to
restrict the typings for collections; and enhancements to polymorphism that
improve precision when analyzing collection operations.

We implemented our type system for Java.  In a case study, it found
previously-unknown nondeterminism errors in a program that had been heavily
vetted by its developers.  In an experiment, it found all of the
non-concurrency-related nondeterminism that was found by other tools.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
When a program is nondeterministic, it is difficult to test and debug.
Nondeterminism can occur even in sequential programs, for example as a
result of iterating over the elements of a hash table. This seemingly innocuous and
frequently used operation can result in diverging test results
if the test depends on iteration order for asserting correctness.

%We have created a type system for determinism.  
We have created a type system that can express determinism specifications
in a program.
A type checker warns if a
program is nondeterministic or verifies that the program is deterministic.
The key ideas in the type system are type qualifiers for nondeterminism,
order-nondeterminism, and determinism; type well-formedness rules to
restrict the typings for collections; and enhancements to polymorphism that
improve precision when analyzing collection operations. While state of the art
nondeterminism detection tools rely on observing runtime output, our approach
aims to verify determinism thereby providing stronger soundness guarantees.

We implemented our type system for Java.  In a case study of a production software that spans
approximately 24K lines of code, it found
previously-unknown nondeterminism errors in a program that had been heavily
vetted by its developers.  In an experiment, it found all of the
non-concurrency-related nondeterminism that was found by other tools.

### 2019-04-06 randoop-case-study.tex

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility.

@@

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop (commits e15f9155--32f72234).
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it (personal communication, 2019).
Some of the nondeterminism was in libraries, such as the JDK\@.
The most common causes were \<toString> routines and iteration order of sets and maps.
The most common fixes were to change the implementation of \<toString>
and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
before iterating over them.
The developers did not make every \<Set> and \<Map> a \<LinkedHashSet> or \<LinkedHashMap>, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
not always desired, and even where it was acceptable, it would have increased costs.

>>>>

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility. \todo{This should include our link to
      our annotated plume-util, which we found was required to reproduce the
    results of the case study}

@@

In July 2017, the Randoop developers spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior in Randoop (commits e15f9155--32f72234).
Their methodology was to repeatedly
 run Randoop with verbose logging enabled,
 look for differences in logging output,
 find the root cause of nondeterminism,
 and eliminate it (personal communication, 2019).
Some of the nondeterminism was in libraries, such as the JDK\@.
The most common causes were \<toString> routines and iteration order of sets and maps.
The most common fixes were to change the implementations of \<toString>
and to use \<LinkedHashSet> and \<LinkedHashMap> or to sort collections
before iterating over them.
The developers did not make every \<Set> and \<Map> a \<LinkedHashSet> or \<LinkedHashMap>, because that
would have increased memory and CPU costs.  They chose not to make every
order-nondeterministic \<List> a \<Set>, for similar reasons:  deduplication was
not always desired, and even where it was acceptable, it would have increased costs.

### 2019-04-06 related.tex

\todo{TODO}

@@

Nondeterminism in tests is of interest to both
researchers and software developers alike~\cite{Fowler,Sudarshan}.
\cite{LuoHEM2014} study 51 open source projects and analyze the most common root causes of
test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
the use of async await, concurrency, or due to test order dependency.~\cite{Plotkin:1993:LPP:645891.671433}
perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
Additionally, the authors of this paper study the effects of removing these test smells. Their results suggest that
refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
these techniques and aims to prevent nondeterminism from causing harmful effects.

The problem of inconsistencies between specifications and implementations
is quite prevalent.~\cite{Jin:2012:UDR:2254064.2254075} study the effect of inefficient code sequences on the performance of software.
Their findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
performance bugs. ~\cite{Rui:2013:180377} analyze the APIs related to security such as authorization and authentication.
These APIs which come as a part of the SDKs provided by online providers are used by application developers to 
build secure apps. The paper discovers that the security guarantees provided by these SDKs are violated due to
the implicit assumptions made by the APIs that the users are unaware of. \cite{Xiao:2014:NMC:2591062.2591177}
study the effect of nondeterminism in MapReduce programs with a specific focus on nondeterminism caused by
non-commutative reducers. While they found bugs that violated correctness due to this bug pattern,
the authors reported that several of these were harmless as they relied on an implicit assumption on data
which ensured correctness.

\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998} propose several techniques to test whether a
deterministic implementation conforms to its non-deterministic finite state machine.
\cite{Cook:2013:RNP:2491956.2491969} present an approach that can automatically verify properties
in branching time temporal logic systems that are inherently non-deterministic.

Failing tests that are unrelated to code changes can be expensive both in terms of monetary costs and
developer effort. \cite{Herzig:2015:EDF:2819009.2819018} propose techniques to classify tests as false alarms
if they are known to be caused by testing infrastructure or other environment issues.~\cite{Huo:2014:IOQ:2635868.2635917}
propose an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigate the effects of test independence assumption
on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}
analyze test dependencies and either prevent them or use this information for other optimizations.
\cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
from those caused by source code. While we haven't fully accounted for all nondeterministic sources due
to the environment, our technique is easily extensible by adding annotations to the corresponding classes.

>>>>

@@

Nondeterminism in tests is of interest to both
researchers and software developers alike~\cite{Fowler,Sudarshan}.
In \cite{LuoHEM2014}, the authors studied 51 open source projects and analyzed the most common root causes of
test flakiness. The findings in the paper suggest that most of the flakiness in tests is caused by
the use of async await, concurrency, or due to test order dependency. The authors of~\cite{Plotkin:1993:LPP:645891.671433}
perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
Additionally,~\cite{Plotkin:1993:LPP:645891.671433} studies the effects of removing these test smells. The results suggest that
refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
these techniques and aims to prevent nondeterminism from causing harmful effects.

The problem of inconsistencies between specifications and implementations
is quite prevalent.~\cite{Jin:2012:UDR:2254064.2254075} looks at the effect of inefficient code sequences on the performance of software.
Its findings suggest that incorrect understanding of performance specifications of APIs is a very common root cause for
performance bugs. ~\cite{Rui:2013:180377} analyzes the APIs related to security such as authorization and authentication.
These APIs which come as a part of the SDKs provided by online providers are used by application developers to 
build secure apps. The paper discovers that the security guarantees provided by these SDKs are violated due to
the implicit assumptions made by the APIs that the users are unaware of. In \cite{Xiao:2014:NMC:2591062.2591177},
the authors
study the effect of nondeterminism in MapReduce programs with a specific focus on nondeterminism caused by
non-commutative reducers. While they found bugs that violated correctness due to this bug pattern,
the authors reported that several of these were harmless as they relied on an implicit assumption on data
which ensured correctness.

Several techniques have been proposed to test whether a
deterministic implementation conforms to its non-deterministic finite state machine~\cite{Petrenko1996,Petrenko:1993:NSM:648128.761244,Savor:1997:639710,Hierons:2004:TCD:1040993.1040998}.
\cite{Cook:2013:RNP:2491956.2491969} presents an approach that can automatically verify properties
in branching time temporal logic systems that are inherently non-deterministic.

Failing tests that are unrelated to code changes can be expensive both in terms of monetary costs and
developer effort. \cite{Herzig:2015:EDF:2819009.2819018} proposes techniques to classify tests as false alarms
if they are known to be caused by testing infrastructure or other environment issues.~\cite{Huo:2014:IOQ:2635868.2635917}
presents an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigates the effects of test independence assumption
on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}
analyze test dependencies and either prevent them or use this information for other optimizations.
The approaches in \cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
from those caused by source code. While we haven't fully accounted for all nondeterministic sources due
to the environment, our technique is easily extensible by adding annotations to the corresponding classes.
